Notes on slotted e-graphs

### 2025-11-12 POPL2026/paper/sections/categorical_semantics.tex
moved all notes to a the same folder

\begin{definition}
	A set of $\Sigma$-terms generated by monoidal signature $\Sigma$ is given by the following grammar:
	\[
		\tau \coloneqq \phi \;|\; \id_{I} \;|\; \id_{A} \;|\; \sym_{A,B} \;|\; \tau_{1};\tau_{2} \;|\; \tau_{1} \otimes \tau_{2},
	\]
	where $\phi$ ranges over the set of operators $\Sigma_{M}$ and $A,B$ range over the set of objects $\Sigma_{O}$.
\end{definition}

@@

\begin{definition}[Symmetric monoidal $\catname{SLat}$-category]%
	\label{def:enriched-prop}
	A symmetric monoidal $\catname{SLat}$-category is given by a $\catname{SLat}$-category $\mathbb{C}$ that additionally has
	\begin{itemize}
		\item a unit object $I \in \mathbb{C}$;
		\item a $\catname{SLat}$-functor tensor $\otimes\colon \mathbb{C} \times \mathbb{C} \to \mathbb{C}$;
	\end{itemize}
	such that particular coherence diagrams analogous to those for symmetric monoidal categories commute.
\end{definition}
This is completely analogous to a symmetric monoidal category, with the fact that the tensor is an $\catname{SLat}$-functor ensuring compatibility with the join: $f \otimes (g+h) = f \otimes g + f \otimes h $ and $ (f+g) \otimes h = f \otimes h + g \otimes h$.
As before, we take $\otimes$ to bind more tightly than $+$.

>>>>

\begin{definition}
	A set of $\Sigma$-terms generated by monoidal signature $\Sigma$ is given by the following grammar:
	$\tau \coloneqq \phi \;|\; \id_{I} \;|\; \id_{A} \;|\; \sym_{A,B} \;|\; \tau_{1};\tau_{2} \;|\; \tau_{1} \otimes \tau_{2}$,
	where $\phi$ ranges over the set of operators $\Sigma_{M}$ and $A,B$ range over the set of objects $\Sigma_{O}$.
\end{definition}

@@

\begin{definition}[Symmetric monoidal $\catname{SLat}$-category]%
	\label{def:enriched-prop}
	A symmetric monoidal $\catname{SLat}$-category is given by a $\catname{SLat}$-category $\mathbb{C}$ that additionally has: a unit object $I \in \mathbb{C}$, an $\catname{SLat}$-functor tensor $\otimes\colon \mathbb{C} \times \mathbb{C} \to \mathbb{C}$, such that particular coherence diagrams analogous to those for symmetric monoidal categories commute.
\end{definition}
This is completely analogous to a symmetric monoidal category, with the fact that the tensor is an $\catname{SLat}$-functor ensuring compatibility with the join: $f \otimes (g+h) = f \otimes g + f \otimes h $ and $ (f+g) \otimes h = f \otimes h + g \otimes h$.
As before, we take $\otimes$ to bind more tightly than $+$.

### 2025-11-12 POPL2026/paper/sections/combinatorial_semantics.tex
moved all notes to a the same folder

To express the rewriting of (hyper-)graphs with interfaces,  i.e.\ the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \emph{interfaces} (DPOI) as shown in~\autoref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish for a rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal{L} \xleftarrow{} m$, $n \xrightarrow{} \mathcal{R} \xleftarrow{} m$.
By noting that a cospan $0 \xrightarrow{} \mathcal{L} \xleftarrow{[f,g]} n \sqcup m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ \cite[Remark 3.14]{bonchi_string_2022-1} we can formulate a rewrite rule as a single span $\mathcal{L} \xleftarrow{} n \sqcup m \xrightarrow{} \mathcal{R}$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal{G}$ and $\mathcal{H}$.
We now make this identification throughout without further mention.
Thus, we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]%
	\label{def:dpoi}
	Given a span of morphisms $\mathcal{G} \xleftarrow{} n \sqcup m \xrightarrow{} \mathcal{H}$ in $\catname{Hyp}(\Sigma)$,  we say $\mathcal{G}$ rewrites to $\mathcal{H}$ with interface $n \sqcup m$ via rewrite rule $\mathcal{L} \xleftarrow{} i \sqcup j \xrightarrow{} \mathcal{R}$ if there exists an object $\mathcal{C}$ and morphisms which complete the rightmost commutative diagram in~\autoref{fig:dpo_dpoi} such that the two marked squares are pushouts.

>>>>

To express the rewriting of (hyper-)graphs with interfaces,  i.e.\ the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \emph{interfaces} (DPOI) as shown in~\autoref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish for a rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal{L} \xleftarrow{} m$, $n \xrightarrow{} \mathcal{R} \xleftarrow{} m$.
By noting that a cospan $0 \xrightarrow{} \mathcal{L} \xleftarrow{[f,g]} n \sqcup m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ \cite[Remark 3.14]{bonchi_string_2022-1} we can formulate a rewrite rule as a single span $\mathcal{L} \xleftarrow{} n \sqcup m \xrightarrow{} \mathcal{R}$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal{G}$ and $\mathcal{H}$.
We now make this identification throughout without further mention.
Thus, we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]%
	\label{def:dpoi}
	Given a span of morphisms $\mathcal{G} \xleftarrow{} n \sqcup m \xrightarrow{} \mathcal{H}$ in $\catname{Hyp}(\Sigma)$,  we say $\mathcal{G}$ rewrites to $\mathcal{H}$ with interface $n \sqcup m$ via rewrite rule $\mathcal{L} \xleftarrow{} i \sqcup j \xrightarrow{} \mathcal{R}$ if there exists an object $\mathcal{L}^{\bot}$ and morphisms which complete the rightmost commutative diagram in~\autoref{fig:dpo_dpoi} such that the two marked squares are pushouts.

### 2025-11-12 notes/SlottedEGraphs/SlottedNotes.tex
moved all notes to a the same folder

>>>>

\subsection{Shape}

A \textit{shape} is a pair $(e, \sigma)$ where $e$ is an e-node and $\sigma$ is a slot map for this particular e-node.
The e-node $e$ can be thought of a canonicalized version of the e-node $\sigma(e)$ (i.e. slot map applied to $e$). The \textit{hashcons} data structure maps such canonicalized e-nodes to e-class ids.
This is how non-duplication is maintained.
Below is the code for computing the shape.
In plain words, it takes an e-node $c$ and replaces all the slots in $c$ with canonical numerical slots $\$1, \$2$ etc..
For example, given an e-node \mintinline{rust}{Var(Slot::named("x"))} (a code for $\texttt{var \$x}$ e-node), the canonicalized version will be $\texttt{var \$1}$ (or, \mintinline{rust}{Var(Slot::numeric(1))}).
Now, to give a more complex example, suppose we were to add $\lambda x . x$ and $\lambda y . y$ expressions to a slotted e-graph.
In code this would be 
\begin{minted}{rust}
  Lambda::Lam(Bind { 
    slot: Slot::named("x"),
    elem: AppliedId { id: Id(0), m: {"f0" -> "x"}}
  })
\end{minted}

Where \mintinline{rust}{AppliedId { id: Id(0), m: {"f0" -> "x"}}} is the e-class id for the variable $x$ inside $\lambda x . x$.
We get the same e-node for the $\lambda y . y$ with "x" replaced by "y".
The slots of this e-node are the two occurrences of slot $x$ (one for the binder and another one from the values of the \mintinline{rust}{AppliedId}).
Therefore, the first component of the shape will be

\begin{minted}{rust}
  Lambda::Lam(Bind { 
    slot: Slot::numeric(1),
    elem: AppliedId { id: Id(0), m: {"f0" -> 1}}
  })
\end{minted}

and we get the same shape for $\lambda y . y$.

\begin{listing}
  \begin{minted}[linenos, fontsize=\small, breaklines]{rust}
    fn weak_shape(&self) -> (Self, Bijection) {
      let mut c = self.clone();
      let mut m = SlotMap::new();
      let mut i = 0;

      for x in c.all_slot_occurrences_mut() {
          let x_val = *x;
          if !m.contains_key(x_val) {
              let new_slot = Slot::numeric(i);
              i += 1;

              m.insert(x_val, new_slot);
          }

          *x = m[x_val];
      }

      let m = m.inverse();

      let public = c.slots();
      let m: SlotMap = m.iter().filter(|(x, _)| public.contains(x)).collect();

      (c, m)
  }
  \end{minted}
\end{listing}
